Liveness-Based Garbage Collection for Lazy Languages

### 2015-06-05 paper/paper.tex
the statement of undecidability

%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
\newcommand{\stk}{\mbox{S}}       % stack
\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\begin{document}
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

The  solution  is  to   also  treat  the  operations  mentioned  above
symbolically.  We  introduce three  new  symbols.   First, \bcar\  and
\bcdr\ are defined as:
\begin{align*}
 \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
 \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
\end{align*}
and \cbar\ as:
\begin{align*}
 \cbar\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
                       & \{\epsilon\} ~\mbox{otherwise}
\end{align*}
Note that \cbar\ is idempotent. 
We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
\begin{align*}
&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
\mbox{and} \\
&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          =   x.\cbar\sigma \cup x.0\sigma
\end{align*}
and the \Lfunonly\ rule
for \SIF\ as:
\begin{align*}
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
        &[\,\mathit{ep} \mapsto  x.\cbar \sigma]
\end{align*}
The rules for \CDR, \PRIM\ and  \NULLQ\  are also modified in a manner
similar to \CAR.
The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{align*}
  &\sigma_1\bcar\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
&\sigma_1\bcdr\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \acdr\sigma_2',~\mbox{and}\\
&\sigma_1\cbar \emptyset \hookrightarrow \emptyset
\end{align*}

The reflexive  transitive closure of  $\hookrightarrow$ will
be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
following proposition relates  the original and the modified
liveness rules.

@@

In general, the equations for  \Lfonly\ are recursive since, as in the
case  of  \length,  \Lan{1}\  may   have  been  defined  in  terms  of
\Lfone{\mathit{f}}.  However,  it is desirable  to have a  closed form
solution for $\Lfone{\mathit{f}}$.  As mentioned in \cite{asati14lgc},
each  of the liveness  rules modifies  a demand  by prefixing  it with
symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \cbar
\rbrace$, and therefore we can assume that $\Lf{f}{i}{\sigma}$ has the
closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}.\sigma
\end{eqnarray}
where \Df{f}{i} are sets of strings over the alphabet mentioned above.
Substituting   the   guessed   form   in   the   equation   describing
\Lfonly$_{\mathit f}$, and factoring  out $\sigma$, we get an equation
for  \Df{f}{i} that  is  independent of  $\sigma$.   Any solution  for
\Df{f}{i}  yields a  solution for  \Lfonly$_{\mathit f}$.   Applied to
\Lfonly$_{\mathit \length}$, we get:
  \begin{eqnarray*}
&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}.\sigma,~\mbox{and}\\
&&   \Df{\length}{1} = \cbar \cup \acdr\Df{\length}{1}\cbar 
       \cup \cbar\Df{\length}{1}.{\cbar}
  \end{eqnarray*}

@@

To avoid analyzing the body of a function for each call, we calculated
the liveness  for the arguments and  the variables in  a function with
respect to a symbolic demand  $\sigma$.  To get the actual liveness we
calculate an over-approximation of the  actual demands made by all the
calls by a {\em summary demand} and calculate the liveness at each GC
point \comment{have we defined what a garbage collection point
  is?} inside the  function based on this summary  demand.  Even though
we  lose precision  due to  this, our  experiments show  that  this is
acceptable  \comment{Is  this  claim  true?}.   The  0-CFA-style  {\em
  summary demand}  is calculated by taking  a union of  the demands at
every call site of a function.

@@

For  the  running  example,   $\length$  has  calls  from  $\main$  at
$\pi_{10}$ and a recursive call at $\pi_6$.
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{11}}{\main}  \cup 
\deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\length}    &=&
 \sigma_{\!all}  ~\cup~{\cbar\sigma_{\mathit{\length}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pb} in terms
of  $\sigma_{\length}$ are:
\begin{align*}
\Lanv{1}{\pl} &= \cbar\sigma_{\length} \cup \acdr\Lf{\length}{1}{\cbar\sigma_{\length}}
  \cup \cbar\Lf{\length}{1}{\cbar\sigma_{\length}}
 \\
\Lanv{9}{\pb} &= \Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}
\end{align*}

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item   For    each   function   $\mathit{f}$,    equations   defining
\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item  For  each   function  $\mathit{f}$, an  equation  defining
the summary demand  $\sigma_{\mathit f}$ on $e_f$.
\item For each function $\mathit{f}$ (including $\mainpgm$ for
  $e_\mainpgm$) 
an equation defining
  liveness at each GC point of $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal
for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{5}{\xl}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{5}{\xl}}  &\rightarrow&   \var{\acdr \Lf
  {\length}{1}{\cbar\sigma_{\length}}} \mid \cbar\sigma_{\length} \\
\end{eqnarray*}
\comment{Pick up a richer equation or more than one liveness equations}
Other equations can be converted  similarly.     The  language
generated  by  \var{\Lanv{i}{x}},
denoted $\mathscr{L}(\var{\Lanv{i}{x}})
$,  is the desired  solution of
\Lanv{i}{x}.
However, recall from  our earlier  discussion that
the  decision problem that  we are  interested in  during garbage
collection is:
\begin{quote}
Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let
$\mathscr{L}(\var{\Lanv{i}{x}})
\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists
of forward paths only. Then does $\alpha \in \sigma$?
\end{quote}
We could convert  the  rules  defining  $\hookrightarrow$  into
productions and  add them to the grammar. However, this results
in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
membership  problem   for  such  grammars   is
undecidable. \comment{Rewrite it to say that this is a CFG with 3
  unrestricted rules and the membership for even this grammar is
undecidable}
\comment{State the undecidability of the decision problem above.}

>>>>

%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
\newcommand{\stk}{\mbox{S}}       % stack
\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
\newcommand{\bang}{\mbox{\sc bang}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{lemma}[theorem]{Lemma}
\begin{document}
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

The  solution  is  to   also  treat  the  operations  mentioned  above
symbolically.  We  introduce three  new  symbols.   First, \bcar\  and
\bcdr\ are defined as:
\begin{align*}
 \bcar\sigma &\triangleq \{\alpha \mid 0\alpha \in \sigma\}\\
 \bcdr\sigma &\triangleq \{\alpha \mid 1\alpha \in \sigma\}
\end{align*}
and \cbar\ as:
\begin{align*}
 \cbar\sigma \triangleq\; & \emptyset~\mbox{if}~\sigma = \emptyset\\
                       & \{\epsilon\} ~\mbox{otherwise}
\end{align*}
Note that \cbar\ is idempotent. 
We can now rewrite the \CONS\ and the \CAR\ rules of $\mathit{ref}$ as:
\begin{align*}
&\mathit{ref\/}((\CONS~x~y),\sigma,\Lfonly)
= x.\bcar\sigma \cup y.\bcdr\sigma  \label{eqn:mod-cons},~
\mbox{and} \\
&\mathit{ref\/}((\CAR~x),\sigma,\Lfonly)
          =   x.\cbar\sigma \cup x.0\sigma
\end{align*}
and the \Lfunonly\ rule
for \SIF\ as:
\begin{align*}
\mathcal{L}((\SIF~x~e_1~e_2),\sigma,\Lfonly) =
                    &\mathcal{L}(e_1,\sigma,\Lfonly)~\uplus
        \mathcal{L}(e_2,\sigma,\Lfonly)~\uplus\\
        &[\,\mathit{ep} \mapsto  x.\cbar \sigma]
\end{align*}
The rules for \CDR, \PRIM\ and  \NULLQ\  are also modified in a manner
similar to \CAR.
The  defintions of \bcar\, \bcdr\ and \cbar\  induce   the  following
relation $\hookrightarrow$ over sets of access paths:
\begin{align*}
  &\sigma_1\bcar\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 = \acar\sigma_2',\\
&\sigma_1\bcdr\sigma_2  \hookrightarrow
  \sigma_1\sigma_2'\mbox{ where } \sigma_2 =
  \acdr\sigma_2',~\mbox{and}\\
&\sigma_1\cbar \sigma_2 \hookrightarrow \epsilon, \mbox{ if } \sigma \ne
  \emptyset \\
&\sigma_1\cbar \emptyset \hookrightarrow \emptyset
\end{align*}
The reflexive  transitive closure of  $\hookrightarrow$ will
be   denoted    as   $\stackrel{*}{\hookrightarrow}$.    The
following proposition relates  the original and the modified
liveness rules.

@@

In general, the equations for  \Lfonly\ are recursive since, as in the
case  of  \length,  \Lan{1}\  may   have  been  defined  in  terms  of
\Lfone{\mathit{f}}.  However,  it is desirable  to have a  closed form
solution for $\Lfone{\mathit{f}}$.  As mentioned in \cite{asati14lgc},
each  of the liveness  rules modifies  a demand  by prefixing  it with
symbols  in the  alphabet  $\lbrace \acar,  \acdr,\bcar, \bcdr,  \cbar
\rbrace$, and therefore we can assume that $\Lf{f}{i}{\sigma}$ has the
closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} = \Df{f}{i}\sigma
\end{eqnarray}
where \Df{f}{i} are sets of strings over the alphabet mentioned above.
Substituting   the   guessed   form   in   the   equation   describing
\Lfonly$_{\mathit f}$, and factoring  out $\sigma$, we get an equation
for  \Df{f}{i} that  is  independent of  $\sigma$.   Any solution  for
\Df{f}{i}  yields a  solution for  \Lfonly$_{\mathit f}$.   Applied to
\Lfonly$_{\mathit \length}$, we get:
  \begin{eqnarray*}
&&  \Lf{\length }{1}{\sigma} = \Df{\length}{1}\sigma,~\mbox{and}\\
&&   \Df{\length}{1} = \cbar \cup \acdr\Df{\length}{1}\cbar 
       \cup \cbar\Df{\length}{1}{\cbar}
  \end{eqnarray*}

@@

To avoid analyzing the body of a function for each call, we calculated
the liveness  for the arguments and  the variables in  a function with
respect to a symbolic demand  $\sigma$.  To get the actual liveness we
calculate an over-approximation of the  actual demands made by all the
calls by a {\em summary demand}  and calculate the liveness at each GC
point inside the  function based on this summary  demand.  Even though
we  lose precision  due to  this, our  experiments show  that  this is
acceptable  \comment{Is  this  claim  true?}.   The  0-CFA-style  {\em
  summary demand}  is calculated by taking  a union of  the demands at
every call site of a function.

@@

For  the  running  example,   $\length$  has  calls  from  $\main$  at
$\pi_{10}$ and a recursive call at $\pi_6$.
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{11}}{\main}  \cup 
\deltacall{\length}{\pi_6}{\length}$.
Filling in  the values gives:
\begin{eqnarray*}
\sigma_{\length}    &=&
 \sigma_{\!all}  ~\cup~{\cbar\sigma_{\mathit{\length}}}
\end{eqnarray*}
As examples, the liveness of \Lanv{1}{\pl} and \Lanv{9}{\pb} in terms
of  $\sigma_{\length}$ are:
\begin{align*}
\Lanv{1}{\pl} &= \cbar\sigma_{\length} \cup \acdr\Df{\length}{1}\cbar\sigma_{\length}
  \cup \cbar\Df{\length}{1}{\cbar\sigma_{\length}}
 \\
\Lanv{9}{\pb} &= \Df{\length}{1}\sigma_{\mathit{\!all}}
\end{align*}

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item   For    each   function   $\mathit{f}$,    equations   defining
\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item  For  each   function  $\mathit{f}$, an  equation  defining
the summary demand  $\sigma_{\mathit f}$ on $e_f$.
\item For each function $\mathit{f}$ (including $\mainpgm$ for
  $e_\mainpgm$) 
an equation defining
  liveness at each GC point of $e_{\mathit f}$.
\end{enumerate}
\subsection{Solving liveness equations---the grammar
interpretation}\label{sec:grammar-formulation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal
for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{1}{\xl}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{1}{\xl}}  &\rightarrow&   \cbar\sigma_{\length} \mid \acdr \var{\Df
  {\length}{1}}{\cbar\sigma_{\length}} \mid
  \cbar\var{\Df{\length}{1}}{\cbar\sigma_{\length}} \\
  \var{\Df{\length}{1}} &\rightarrow& \cbar \mid
  \acdr\var{\Df{\length}{1}}\cbar 
       \mid \cbar\var{\Df{\length}{1}}{\cbar}\\
\langle {\sigma_{\length}} \rangle
&\rightarrow&
\sigma_{\!all}  \mid \cbar\langle{\sigma_{\length}}\rangle
\end{eqnarray*}
Other equations can be converted similarly.  The language generated by
\var{\Lanv{i}{x}},  denoted $\mathscr{L}(\var{\Lanv{i}{x}}) $,  is the
desired  solution of  \Lanv{i}{x}.  However,  recall from  our earlier
discussion that the decision problem  that we are interested in during
garbage collection at a program point $\pi_i$ is:
\begin{quote}
Let  $x.\alpha$ be  a forward  access  path consisting  only of  edges
\acar\    and   \acdr\    (but    not   \bcar, \bcdr\ or \cbar).     Let
$\mathscr{L}(\var{\Lanv{i}{x}}) \stackrel{*}{\hookrightarrow} \sigma$,
where $\sigma$ consists  of forward paths only. Then  does $\alpha \in
\sigma$?
\end{quote}

We now model the above problem  as one of deciding the membership of a
context-free  grammar  augmented  with  a fixed  set  of  unrestricted
productions.

\begin{definition}\label{def:specialgrammar} 
Consider a form of  grammar $(N,T, p_1\cup p_2,S)$ in which $N$ is a set of
non-terminals, $T = \{\acar, \acdr, \bcar, \bcdr, \cbar, \$\}$, $p_1$
is a set of context-free production that contains the distinguished
production $S \rightarrow \alpha\$$, $\alpha$ is a string of grammar
symbols that does not contain $S$, and $p_2$ is the fixed set of
unrestricted productions $\bcar\acar \rightarrow \epsilon$,
$\bcdr\acdr \rightarrow \epsilon$, $\cbar\acar \rightarrow \cbar$,
$\cbar\acdr \rightarrow \cbar$, and $\cbar\$ \rightarrow
\epsilon$. 
\end{definition}

From Sections \ref{sec:liveness}  and \ref{sec:computing}, it is clear
that the results of liveness analysis of any program can be modeled by
the kind  of grammar described above. The  following proposition shows
that the converse also holds.
\begin{proposition}
Given    a     grammar    $G$    of    the     form    described    in
Definition~\ref{def:specialgrammar},  it is  possible  to construct  a
program $p$  with program points  $\pi_i$ and variables $x$  such that
the liveness  analysis of $p$  is same as  $G$ except for a  change in
non-terminal names which now have the form \var{\Lanv{i}{x}}.
\end{proposition}
We now show that the decision problem required at garbage collection
time is undecidable.

\begin{lemma}
Consider a grammar $G$ of the kind  described in
Definition~\ref{def:specialgrammar} and a forward access path $\alpha$ consisting
of symbols \acar\ and \acdr\  only. The decision problem $\alpha \in
\mathscr{L}(G)$ is undecidable.  
\end{lemma}

\comment{Proof:}

%% We could convert the rules defining $\hookrightarrow$ into productions
%% and  add  them  to the  grammar.  However,  this  results in  an  {\em
%%   unrestricted   grammar}~\cite{hopcraft90toc},  and   the  membership
%% problem for  such grammars is undecidable. \comment{Rewrite  it to say
%%   that this is a CFG with  3 unrestricted rules and the membership for
%%   even this grammar  is undecidable} \comment{State the undecidability
%%   of the decision problem above.}
